Dancing on Water: The Choreography of Sulfur Dioxide Adsorption to Aqueous Surfaces

### 2011-10-25 resubmission/Article.tex
Final resubmission to the journal

\usepackage {amsmath}
\usepackage {setspace}
\usepackage [pdftex]{graphicx}
\usepackage[sort&compress]{achemso}

\oddsidemargin  0.0in
\evensidemargin 0.0in
\textwidth      6.5in

@@

\doublespacing

@@

Molecular dynamics simulations were performed using the Amber 11 software suite.\cite{Case2010} Polarizable models for the \wat~and \suldiox~molecules were used in the simulations, and have been used previously in studies on interfacial systems because they are known to more accurately reproduce interfacial structure and free energy profiles.\cite{Wick2007,Rivera2006,Dang1998} The \wat~model used is the POL3 water model,\cite{Caldwell1995} and for \suldiox~we used the model of Baer et al. that places a single polarizable center on the sulfur atom.\cite{Baer2010}

@@

The low and high concentration simulated systems, ``neat-water'' and ``saturated water'', respectively, were created as follows: the neat-water simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true neat-water air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water. Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an added \suldiox~gas pressure. The saturated system with both bulk and gaseous \suldiox~was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

@@

A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the neat-water and saturated starting systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were then evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfully model chemical events.\cite{Isralewitz2001,Giorgino2011,Bizzarri2011,Strzelecki2009,Patargias2009,Liu2006} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a time step of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

\subsection {Water Surface Location}

@@

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation. Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin gray line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted Gibbs dividing surface has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{Humphrey1996} Alternatively, assuming a spherical model of water with a radius of 2.2\angs~(the typical length of a bulk-water hydrogen bond), two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

@@

@@

	The ``tilt'' angle, $\theta$, defines the angle formed between the molecular bisector vector (the molecular z-axis, pointing from the central atom in the direction of the other two atoms) and the positive system reference axis. Thus the value of $\theta$ falls within a range of $[0,\pi]$. An angle of $\cos(\theta)=1$ indicates a molecule with its bisector aligned with the reference axis, while $\cos(\theta)=-1$ results from an anti-aligned configuration. Sample representations of different values of $\cos(\theta)$ are shown in Figure \ref{fig:water-angles}B.

	A second angle, $\phi$, defines the molecular ``twist'' for certain configurations of the molecule. $\phi$ is the angle formed between the vector normal to the plane of the molecule (molecular y-axis) and the system reference axis. The values of $\phi$ fall in the interval $[0,\frac{\pi}{2}]$ because of the symmetry of \wat~and \suldiox~molecules with respect to twist about their bisector axes. For values of $\cos(\theta) \approx 0$, $\phi$ provides additional information about whether the molecular orientation is ``flat'' to the surface (e.g. the plane of the molecule is aligned with the plane of the surface), or if it is perpendicular. The values of $\cos(\phi)$ for different molecular orientations are shown in Figure \ref{fig:water-angles}C. %However, the symmetry of the \wat~and \suldiox~molecules create equivalence between the two H, or O, atoms (in the \wat~or\suldiox, respectively). $\phi=-\pi$ is equivalent to $\phi=\pi$, and both situations are equivalent to $\phi=0$. Because $\phi=-\phi$, the value is reported within the range of $[0,\frac \pi 2]$. 
	However, values of $\cos(\theta)$ close to 1 or -1 result in an isotropic distribution in $\cos(\phi)$ because of the symmetry of the plane of the surface in directions perpendicular to the reference axis.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{molecular-angles.png}
		\caption{(A) The definition of the angles $\theta$ and $\phi$ used to define molecular orientation of \suldiox~and \wat~relative to a reference axis. (B) $\theta$ is the value of the ``tilt'' of the molecular bisector from the reference axis, with $\cos(\theta)$ values ranging from -1 to +1 indicating different bisector orientations. (C) $\phi$ is the ``twist'' angle formed between the vector normal to the molecular plane (the body-fixed y-axis) and the reference axis. For values of $\cos(\theta) \approx 0$, $\phi$ shows how flat the molecule is relative to the interfacial plane.}
		\label{fig:water-angles}
	\end{center}
\end{figure}

@@

One measure of surface activity is the spatial distribution of molecules in the interfacial region. The density distributions of both \wat~and \suldiox~were calculated for the equilibrium MD simulations. The results presented in Figure \ref{fig:density} show both the water (black) and \suldiox~(red) density distributions across the two simulated slabs. As shown, the single \suldiox~in the neat-water system remained at the water slab surface. The \suldiox~in the saturated system accumulated mostly at the surface, but some residual \suldiox~remained well into the bulk water.

@@

Geometric analyses were performed to characterize the net molecular orientation of \wat~and \suldiox~molecules at different depths from the water surface location. At each distance from the surface location, an orientation profile was created for both the \wat~and \suldiox~molecules. The orientation distribution for the angles $\theta$ and $\phi$ at all depths were combined to form 2D intensity plots that show how the molecular orientation distributions change with distance to the surface location. These plots (Figures \ref{fig:water-orientation}, \ref{fig:so2-transit-angles}, and \ref{fig:so2-surface-angles}) allow for a visual interpretation of how the net orientations are affected when moving from the gas phase through the interfacial region and to the surface location, and then further into the aqueous interfacial region and bulk. Both the neat-water, with only a single \suldiox~introduced, and the high-concentration saturated system were analyzed. In the case of the neat-water system, the introduction of a single \suldiox~does not greatly affect molecular orientation of water molecules in the interfacial region. These results of the water orientation are very similar to a neat-water system without any adsorbed solutes (not shown).

The 2D orientational depth profile plots show areas of low intensity in dark blue, and highest intensity in dark red. Regions of the plots where the intensity (coloration) is equally distributed from top to bottom along the entire orientational range are considered isotropic. Likewise, areas of the plot with high intensity over a small orientational range are considered to exhibit an orientational preference at the given depth.  The histograms are arranged with the plots of $\cos(\theta)$ on the left and $\cos(\phi)$ on the right. The surface is located at 0\angs. The angle distributions from both simulated slab surfaces were averaged for all the orientation analyses.

@@

The orientation depth-profiles for \wat~are shown in Figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. The interfacial region for both these calculations and the VSF experiments is defined as the region where molecular orientational anisotropy exists around the surface water location. Referring to Figure \ref{fig:water-orientation}, calculations indicate an interfacial width of approximately 10\angs~for the neat-water system, and approximately 16 \angs~for the saturated system. In both systems the strongest orientational preference is found at the slab surfaces (positions above 0\angs) where the water is furthest towards the gas phase. Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The narrow "peninsula" of high intensity that extends in the $\cos(\theta)$ plots to the right of the -5\angs~location shows the overall preference of water to orient at the surface. The $\cos(\phi)$ plots are similar to each other with a narrow region of reorientation, but the effect in the interfacial region is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change.

The bisector tilt of the water molecules, $\cos(\theta)$, concentrates around $\cos(\theta)=0$ within the first few\angs~above and below the water surface location, becoming progressively isotropic further through the interfacial region and into the water bulk of both systems. As the tilt nears $\cos(\theta)=0$ the \wat~bisector lies within the plane of the surface indicating a water orientation either flat on the surface, or with some amount of ``twist'' sending the OH bonds in towards, or out of the bulk. The value of $\phi$ determines the ``twist'' in this case. Both systems show a defined intensity concentration (dark red) in the distributions around $\cos(\phi)=1$ on the aqueous side of the interfacial region. This results from an orientation of the water's y-axis (normal to the molecular plane) aligned perpendicular to the plane of the water surface. Thus, for both the neat-water and saturated systems, the preferred orientation of water molecules at a distance of 0\angs~ is to lie mostly flat to the plane of the interface with a slight ``twist'' sending one OH bond further into the gas phase, and the other OH towards the water bulk. This result agrees somewhat with a recent air/water study by Fan et al. in which the surface orientations of several water models were analyzed.\cite{Fan2009} They simulated water using non-polarizable models, however, which alters the behavior of water at interfacial regions when compared to the polarizable POL3 model used in our work. The main conclusions are similar in that one OH tends to point further into the air phase than the other, but differs in that this effect is less pronounced with the polarizable model.

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the orientation of those waters furthest into the gas phase. For the saturated solution the resulting orientation of waters above 0\angs, shown in the saturated plot of $\cos(\theta)$ of Figure \ref{fig:water-orientation}, is with a bisector pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced further from the water phase, and above 5\angs~the $\cos(\theta)$ distribution is completely centered around $\cos(\theta)=+1$ (see Figure \ref{fig:water-angles}).

The noise in the neat-water plots of Figure \ref{fig:water-orientation} above 5\angs~(manifested as disconnected points of intensity throughout the range of orientations) is a result of fewer waters venturing beyond those extents and thus less data far from the water surface location. This is one indication that waters near a layer of adsorbed \suldiox~venture further above the water surface location relative to the low \suldiox~concentration, where they can have more \suldiox~interaction with the adsorbed gas molecules. These results show that the interactions with neighboring \suldiox~molecules allow the waters above the water surface location to orient more perpendicularly to the interface. This is consistent with our recent experimental VSFS studies which showed evidence for the reorienting behavior of water due to the \suldiox~interactions with the topmost surface waters.\cite{Ota2011}

The distribution of $\cos(\phi)$ is more sharply defined (i.e. less isotropic) for the neat-water system than for the saturated one. Waters on the neat surface lie flatter, whereas the presence of the \suldiox~allows a greater range of ``twist'' for those waters in the plane of the interface. The $\phi$ distributions quickly become isotropic above the surface location as the bisectors orient more perpendicularly, and below the surface as the bulk water loses any orientational preference.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{h2oangles.png}
		\caption{Molecular orientation histograms of \wat~throughout the surface equilibrated systems. The top surface is located at a distance of 0\angs~with negative distance values located in the bulk of the slab. Shown are the angle distributions for $\theta$ (left column) and $\phi$ (right column) in both the neat-\wat~system (top row) and the saturated system (bottom row). The distributions are normalized to account for the changing number of water molecules at different positions in the system. Regions of high intensity are dark red, and low intensity are dark blue. The scattered points of coloration to the far right of each plot indicates that few waters were located at those positions, and thus few data points were collected.}
		\label{fig:water-orientation}
	\end{center}
\end{figure}

@@

Orientation distributions of the adsorbed \suldiox~molecules were created during the equilibrium simulations for both the neat-water and saturated systems. Figure \ref{fig:so2-orientation} shows the distributions of $\cos(\theta)$ and $\cos(\phi)$ (arranged similarly to the water orientation distributions plots in Figure \ref{fig:water-orientation}). The \suldiox~orientation data set for the neat-water system is much smaller as only a single \suldiox~molecule was simulated in the bulk. The resulting distribution plots are thus noisier than the corresponding saturated plots with more scattered points of high intensity, but effective comparisons can still be drawn.

>>>>

\usepackage {amsmath}
\usepackage {textcomp}
\usepackage {setspace}
\usepackage [pdftex]{graphicx}
\usepackage[sort&compress]{achemso}
%\usepackage{lineno}

\usepackage[
top    = 0.5in,
bottom = 1.0in]{geometry}
%left   = 2.50cm,
%right  = 2.50cm]{geometry}

% some formatting tags
%\oddsidemargin  0.0in
%\evensidemargin 0.0in
%\textwidth      6.5in

@@

%\onehalfspacing
%\linenumbers 
\doublespacing

\begin{abstract}
One might expect the high surface tension of water to be a barrier to absorption of a gas into the liquid phase, but we know that gaseous adsorption onto, and subsequent absorption into a water surface is a common phenomenon on this planet.  What is not commonly known is how an atmospheric gas such as \suldiox~and molecules at the water surface can overcome the barrier created by strong water-water surface bonding interactions.  What this interplay looks like, the distances from the water surface at which these attractive interactions begin, and how they influence the orientational nature of both \suldiox~and surface water molecules is the focus of this computational study.  The results fill a void in the information about this system existing from previous experimental studies by providing information about the dimensional nature of the gas-surface interactions, and the details of how the two species twist and turn orientationally with increased surface interactions.  Classical molecular dynamics have been employed in both equilibrium and steered molecular dynamics (SMD) simulations for \suldiox~at a neat-water surface, and at a surface with high interfacial \suldiox~concentrations.  The results provide new molecular insights for understanding the interaction of this prevalent gas on aerosols and other aqueous surfaces in the environment.
\end{abstract}

@@

Molecular dynamics simulations were performed using the Amber 11 software suite.\cite{Case2010} Polarizable models for the \wat~and \suldiox~molecules were used in the simulations, and have been used previously in studies on interfacial systems because they are known to more accurately reproduce interfacial structure and free energy profiles.\cite{Wick2007,Rivera2006,Dang1998} The \wat~model used is the POL3 water model (also discussed in the supplemental information),\cite{Caldwell1995} and for \suldiox~we used the model of Baer et al. that places a single polarizable center on the sulfur atom.\cite{Baer2010} An intermolecular cutoff of 12\angs~was used for long-range electrostatic forces. The simulations were performed in the NVT ensemble using Langevin dynamics for temperature control. Induced dipoles were treated by the polarizable potential functions of the Amber molecular dynamics software.

@@

The low and high concentration simulated systems, ``neat-water'' and ``saturated'', respectively, were created as follows: the neat-water simulation involved the addition of a single \suldiox~molecule either within the bulk of the water slab (for equilibrium MD), or above the slab surface (in the SMD simulations). The single-\suldiox~neat-water system was then evolved for 2 ns to produce an equilibrated starting configuration. Because of the extremely low concentration of \suldiox~in the neat-water system the surface waters behave similarly to a true neat-water air-liquid interface. The neat-water system orientational results shown later in this work reproduce well the results of our previous orientational studies of surface water behavior.\cite{Walker2006b,Hore2008} The saturated system had 22 \suldiox~molecules introduced to the water slab bulk in order to saturate it to a level coinciding with the Henry's law constant for \suldiox~in water (k\textdegree$_H = 1.4$ mol/kg*bar).\cite{Lide2000} Additionally, 50 \suldiox~molecules were introduced into the gas phase outside of the saturated water slab to simulate an added \suldiox~gas pressure. The additional gas in the vapor phase was added over the course of several ns to keep a constant 1 atm of \suldiox~pressure above the water surface as \suldiox~gas molecules adsorbed to the surface. The saturated system with both bulk and gaseous \suldiox~was then evolved for 2 ns to produce a starting configuration for further saturated simulations.

@@

% need to add in the specifics about the SMD routine, force constants, calculations, etc.
A second set of simulations began with an equilibrated water slab as in the surface equilibrated method above. However, in both the neat-water and saturated starting systems, a single \suldiox~was introduced 20\angs~above the water slab surface, with the sulfur atom tethered to its initial position. The systems were then evolved for 1 ns, taking coordinate snapshots every 20 ps to create 50 starting points for further simulations. Steered molecular dynamics (SMD) were then performed on the 50 system configurations (in both the neat-water and saturated configurations) to guide the \suldiox~down towards a tethered water near the water slab's center of mass by applying a small steering force to the \suldiox-sulfur atom. This steering technique has been previously developed and used to successfully model chemical events.\cite{Isralewitz2001,Giorgino2011,Bizzarri2011,Strzelecki2009,Patargias2009,Liu2006} The \suldiox~thus passed through the continuum of environments from gas phase to (neat- and saturated) water surface adsorption, and finally absorption into the bulk of the \wat~slab. Each of the SMD simulations were performed for a total of 200 ps, using a time step of 1 fs, and taking snapshots of the system every 25 fs. Figure \ref{fig:starting-configurations} illustrates two sample starting configurations for the SMD simulations, showing both the neat-water slab and the saturated slab configurations before steering the \suldiox~towards the water bulk.

@@

\subsection {Aqueous Surface Location}

@@

Figure \ref{fig:density-flaw} demonstrates the problem of surface location drift during a simulation (even with utilizing the Amber NSCM parameter). Figure \ref{fig:density-flaw}A shows the density profile of water and \suldiox~over the course of one of the 10 ns trajectories used in this work using the original uncorrected coordinates of the system taken from the raw atomic positions of the molecular dynamics data output. The water density profile and location (thin gray line in Figure \ref{fig:density-flaw}A) is produced by averaging the instantaneous density profile at each time step in the simulation over all the time steps. The water profile was then fit to a $\tanh$ function (black line) to extract the position and width parameters of the water surface. The fitted surface water density profile has a width of 3.77\angs, which is comparable to values reported for similar neat-water systems.\cite{Dang1997,Hore2008} A bulge in the gas-phase ($\leq 52$\angs) side of the water density profile is indicative of the drift of the water slab over the course of the trajectory. Thus the calculated location and width from the $\tanh$ line fit are not accurate over long trajectories for defining a stationary reference point.

	To overcome this problem we define the water surface location by calculating a reference location at each time step by averaging the positions of the waters contained in the topmost monolayer. This provides a consistent and intuitive reference point in the simulations to which we relate our analyses, but does not increase the computational burden. The number of waters included in the averaging is determined by taking a few issues into account. First, counting the waters found in the topmost cross-section of the water slab over several time steps indicated between 65-75 waters that established a full monolayer. This was done by a visual inspection of the slab using the VMD MD visualization package.\cite{Humphrey1996} Alternatively, assuming a spherical model of water with a radius of 2.2\angs, two layers of hexagonally tight-packed spheres yielded a similar number of surface water molecules. Increasing the number of waters used in calculating the surface location diminishes the effects of the few waters that briefly rise above the surface into the gas phase, stabilizing both the surface position and thickness values. Taking the close-packed model as a maximum number of waters fit into a flat surface, the topmost 70 water molecules were used for calculating instantaneous water surface locations for each simulation step.

@@

  %Taking one of the simulations of the neat-water slabs, a series of calculations were performed to find the averaged positions and the standard deviations of varying numbers of the topmost waters. It was found that using 70 waters produced a standard deviation of the water positions corresponding to a width of 3.2\angs. as reported by Baer et al.when fitting the density profile with a hyperbolic tangent lineshape. To match the interfacial width of Baer using the standard deviation of our simulations, 70 water molecules would have to be included in the top monolayer. Thus, the calculations and analyses presented in this work incorporate 70 water molecules when calculating the surface location.

@@

	The ``tilt'' angle, $\theta$, defines the angle formed between the molecular bisector vector (the molecular z-axis, pointing from the central atom in the direction of the other two atoms) and the positive system reference axis. Thus the value of $\theta$ falls within a range of [0\textdegree,180\textdegree]. An angle of $\theta=$ 0\textdegree~indicates a molecule with its bisector aligned with the reference axis, while $\theta=$ 180\textdegree~results from an anti-aligned configuration. Sample representations of molecular orientations resulting from different values of $\theta$ are shown in Figure \ref{fig:water-angles}B.

	A second angle, $\phi$, defines the molecular ``twist'' of the molecule. $\phi$ is the angle of rotation around the molecular bisector axis that quantifies the rotation of the molecular plane with respect to a plane perpendicular to the water surface. The values of $\phi$ fall in the interval [0\textdegree,90\textdegree] because of the symmetry of \wat~and \suldiox~molecules with respect to twist about their bisector axes. For values of $\theta \approx$ 90\textdegree, $\phi$ provides additional information about whether the molecular orientation is ``flat'' to the surface (e.g. the plane of the molecule is aligned with the plane of the surface), or if it is perpendicular. The values of $\phi$ for different molecular orientations are depicted in Figure \ref{fig:water-angles}C. %However, the symmetry of the \wat~and \suldiox~molecules create equivalence between the two H, or O, atoms (in the \wat~or\suldiox, respectively). $\phi=-\pi$ is equivalent to $\phi=\pi$, and both situations are equivalent to $\phi=0$. Because $\phi=-\phi$, the value is reported within the range of $[0,\frac \pi 2]$. 
	Values of $\theta$ close to 0\textdegree~or 180\textdegree~result in an isotropic distribution in $\phi$ because of the symmetry of the plane of the surface in directions perpendicular to the surface normal reference axis.

%	The two molecules studied, \wat~and \suldiox, and similarly shaped with a C2$_v$ axis along their bisectors, and a molecular plane is defined by their three atoms. The orientational analyses presented herein focus on two angles used to define the molecular orientation in space. In each analysis a reference axis is used to define molecular angles, and is the z-axis unless otherwise specified. A ``tilt'' angle, $\theta$, defines the angle formed between a vector (generally the bisector vector pointing from the central atom in the direction of the other two atoms) and the positive reference axis. Thus the value of $\theta$ falls within a range of $[0,\pi]$. A second angle, $\phi$, defines the molecular ``twist'' around the reference axis. $\phi$ is formed between the projection of a vector onto the x-y reference plane, and the x-axis. $\phi$ is thus defined relative to the x-axis, and its values are in the interval $[-\pi,\pi]$. However, because of the symmetry of the \wat~and \suldiox~molecules and the equivalence of the two H or O atoms (in the \wat~or\suldiox, respectively), $\phi=-\pi$ is equivalent to $\phi=\pi$, and both situations are equivalent to $\phi=0$. Because $\phi=-\phi$, the value is reported within the range of $[0,\frac \pi 2]$. Figure \ref{fig:spherical-angle} shows the angle definitions using the z-axis as the reference axis.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{molecular-angles-small.png}
		\caption{(A) The definition of the angles $\theta$ and $\phi$ used to define molecular orientation of \suldiox~and \wat~relative to a reference axis. (B) $\theta$ is the value of the ``tilt'' of the molecular bisector from the reference axis, with $\theta$ values ranging from  0\textdegree~(aligned) to 180\textdegree~(anti-aligned). $\theta = $ 90\textdegree~aligns the bisector parallel to the plane of the water surface. Several bisector orientations are shown depicting different $\theta$ values. (C) $\phi$ is the ``twist'' angle defined as a rotation around the molecular bisector axis. A value of 0\textdegree~aligns the plane of the molecule perpendicular to the plane of the water surface. A value of 90\textdegree~rotates the molecule to lie flat in the plane of the water surface. Because of the symmetries of \suldiox, \wat, and the plane of the water surface, the range of $\phi$ values is limited to 0\textdegree~$\leq \phi \leq$~90\textdegree.}
		\label{fig:water-angles}
	\end{center}
\end{figure}

@@

One measure of surface activity is the spatial distribution of molecules in the interfacial region. The density distributions of both \wat~and \suldiox~were calculated for the equilibrium MD simulations. The results presented in Figure \ref{fig:density} show both the water (black) and \suldiox~(red) density distributions, averaged over the two simulated interfaces of each slab. As shown, the single \suldiox~in the neat-water system remained at the water slab surface. The \suldiox~in the saturated system accumulated mostly at the surface, but some residual \suldiox~remained well into the bulk water.

%The density distributions of the \suldiox~were fit to gaussian lineshapes as shown in equation \ref{eq:gaussian} using the common definitions of the variables $A$, $\mu$, and $\sigma$, and given as a function of the distance to the top surface, $z$.

%\begin{equation}
%  \rho(z)=\frac{A}{\sqrt{2 \pi \sigma^2}} e^{-\frac{(z-\mu)^2}{2\sigma^2}}
%  \label{eq:gaussian}
%\end{equation}

@@

Geometric analyses were performed to characterize the net molecular orientation of \wat~and \suldiox~molecules at different depths from the water surface location. At each distance from the surface location, an orientation profile was created for both the \wat~and \suldiox~molecules. The bi-variate orientation distributions for the angles $\theta$ and $\phi$ at various depths were combined to form the intensity plots that show how the molecular orientation distributions change with distance to the surface location. These plots (Figures \ref{fig:water-orientation}, \ref{fig:so2-orientation}, and \ref{fig:so2-transit-angles}) allow for a visual interpretation of how the net orientations are affected when moving from the gas phase through the interfacial region and to the surface location, and then further into the aqueous interfacial region and bulk. Both the neat-water, with only a single \suldiox~introduced, and the high-concentration saturated system were analyzed. In the case of the neat-water system, the introduction of a single \suldiox~does not greatly affect molecular orientation of water molecules in the interfacial region. These results of the water orientation are very similar to a neat-water system without any adsorbed solutes (not shown).

The depth profile plots are arranged as a grid of 2D histograms. Each histogram is calculated for all molecules falling within a particular depth in the water interfacial region. The depth of each plot is marked (in \ang) in the upper-right, with the water surface location set at 0\angs, positive depths lie on the gas-phase side of the surface, and negative depths are on the bulk water side. The horizontal axes of each histogram represent $\theta$ values, and the vertical axes represent $\phi$ values. Areas of low intensity appear in dark blue, and highest intensity in dark red. Regions of the plots where the intensity (coloration) is equally distributed along either the vertical or horizontal axes are considered isotropic in $\phi$ or $\theta$, respectively. Likewise, areas of the plot with high intensity over a small orientational range are considered to exhibit an orientational preference at the given depth. The angle distributions from both simulated slab surfaces were averaged for all the orientation analyses.

@@

The orientation depth-profiles for \wat~are shown in Figure \ref{fig:water-orientation} for both the neat-water (top) and saturated (bottom) systems during the equilibrium MD simulations. The interfacial region for both these calculations and the VSF experiments is defined as the region where molecular orientational anisotropy exists around the surface water location. Fitting the water density profiles we have calculated an interfacial width of approximately 10\angs~for the neat-water system, and approximately 16\angs~for the saturated system. In both systems the strongest orientational preference is found at the slab surfaces (positions near 0\angs). Previous work on orientational preference of water at air surfaces shows the same trend as our neat-water results.\cite{Walker2006b,Hore2008} The swaths of coloration indicating high intensity appearing in the neat-water plots from -4 to +4\angs, (and the corresponding regions in most of the saturated system plots) show the overall preference of water to orient at the surface. The plots of the neat-water and saturated systems are similar to each other with a narrow region of reorientation, but the effect in the interfacial region is greater in the neat-water system as evidenced by the sharper transition in intensity from blue to red, compared to the saturated system that has a less pronounced intensity change over larger areas of the histograms.

The bisector tilt of the water molecules, $\theta$, concentrates around $\theta=$ 90\textdegree within the first few\angs~above and below the water surface location, becoming progressively isotropic further through the interfacial region and into the water bulk of both systems. Above the surface location at positive distances, $\theta \leq$ 90\textdegree~indicates that the water hydrogens tend to point towards the gas-phase side of the surface. As the tilt nears $\theta=$ 90\textdegree~the \wat~bisector lies within the plane of the surface indicating a water orientation either flat on the surface, or with some amount of ``twist'' sending the OH bonds in towards, or out of the bulk. The value of $\phi$ determines the ``twist'' in this case. Both systems show a similar trend where waters at or just below the surface have values of $\phi$ near 90\textdegree, and waters above the surface take on values of $\phi$ near 0\textdegree. This jump in the angular distribution of $\phi$ indicates that waters at or below the surface lie mostly flat in the plane of the interface, and as they move above the surface towards the gas phase, they reorient with one OH bond pointing towards the water bulk, and one pointing out of the surface into the gas. This behavior is more strongly pronounced in the neat-water system where most of the surface waters are not interacting with an adsorbed layer of \suldiox~molecules.

%This result agrees somewhat with a recent air/water study by Fan et al. in which the surface orientations of several water models were analyzed.\cite{Fan2009} They simulated water using non-polarizable models, however, which alters the behavior of water at interfacial regions when compared to the polarizable POL3 model used in our work. The main conclusions are similar in that one OH tends to point further into the air phase than the other, but differs in that this effect is less pronounced with the polarizable model.

Although the plots show overall similarities for both the neat-water and saturated systems, the presence of a layer of adsorbed \suldiox~molecules alters the orientation of those waters furthest into the gas phase. For the saturated solution, the resulting orientation of waters above 0\angs, shown in the bottom set of plots of Figure \ref{fig:water-orientation}, is nearly isotropic in $\phi$, and with $\theta \leq$ 90\textdegree. This results from waters with bisectors pointing further into the adsorbed \suldiox~gas layer, and both hydrogens pointing outward from the aqueous bulk. The effect is more pronounced as the waters move further from the water surface, and above 4\angs~the $\theta$ distribution is mostly concentrated around $\theta=$ 0\textdegree (see Figure \ref{fig:water-angles}).

%The histograms show overall similarities for both systems in their shapes and intensities from approximately 0\angs~and below. The presence of a saturated layer of adsorbed \suldiox~molecules, however, alters the water orientation, but not necessarily the orientational depth of the interface. In both $\theta$ distributions, the orientations of waters in the bulk region are isotropic until 5\angs~below the surface location. At 5\angs~below the surface the water molecules begin to orient with their bisectors within the plane of the interface (perpendicular to the surface normal, $\cos(\theta)\approx 0$). The corresponding location in the plot of $\phi$ shows that those molecules are also mostly flat to the surface ($\cos(\phi)\approx 1$). Moving further out from the bulk and into the gas phase, the distributions show $\cos(\theta)$ increasing. Waters further out from the bulk have fewer bonding interactions, and orient with their hydrogens more towards the gas phase. The bisector pointing further into the gas phase leads to isotropy in the values of $\phi$.

The angle distributions above 6\angs~in the neat-water plots of Figure \ref{fig:water-orientation} are mostly isotropic (manifested as uniform coloration throughout the range of orientations). Furthermore, there are few data points that make up the histograms, a result of fewer waters venturing beyond those extents. Conversely, waters near a layer of adsorbed \suldiox~venture further above the water surface location relative to the low \suldiox~concentration, where they can have interactions with the adsorbed \suldiox~gas molecules. The waters above the water surface location orient perpendicularly to the interface. This is consistent with our recent experimental VSFS studies which showed evidence for the reorienting behavior of water due to the \suldiox~interactions with the topmost surface waters.\cite{Ota2011}

The distribution of $\phi$ is more sharply defined (i.e. less isotropic) for the neat-water system than for the saturated one. Waters on the neat surface lie flat or perpendicular to the surface if they are below or above the water surface location, respectively. The presence of the \suldiox~allows a greater range of ``twist'' for those waters in the plane of the interface. The $\phi$ distributions quickly become isotropic above the saturated water surface location, shown as a uniform coloration across $\phi$ for most values of $\theta$.

%Peaks in the distributions of the neat-\wat~system are more clearly pronounced as their intensities are more concentrated and larger than the surrounding area of the profiles. This difference indicates that the transition from the preferred orientation at the water surface has a sharper distinction from the isotropic bulk than in the system with the saturated \suldiox~surface. It appears that the same orientation trend is present in both systems, but the presence of the \suldiox~at the water surface decreases the degree of water orientation at the interface.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{theta-phi-h2o.png}
		\caption{Molecular orientation histograms of \wat~throughout the surface equilibrated systems. The depth (in \angs) of each histogram within the water interface is numbered in the upper-right of each respective plot. The water surface location is set at a position of 0\angs~with negative distance values located in the bulk of the slab, and positive distances towards the gas phase. Each histogram is a bi-variate angle distributions for $\theta$ (horizontal axes) and $\phi$ (vertical axes) in both the neat-\wat~system (top) and the saturated system (bottom). Regions of high intensity are dark red, and low intensity are dark blue.}
		\label{fig:water-orientation}
	\end{center}
\end{figure}

@@

Orientation distributions of the adsorbed \suldiox~molecules were created during the equilibrium simulations for both the neat-water and saturated systems. Figure \ref{fig:so2-orientation} shows the 2D distributions of $\theta$ and $\phi$ (arranged similarly to the water orientation distributions plots in Figure \ref{fig:water-orientation}). The \suldiox~orientation data set for the neat-water system is much smaller as only a single \suldiox~molecule was simulated in the bulk. The resulting distribution plots are thus representative of the single surface active \suldiox~molecule. The neat-water \suldiox~molecule remains within a narrower region of the interface than the saturated system \suldiox, but effective comparisons can still be drawn. Note that the depth range of the plots in Figure \ref{fig:so2-orientation} is different for the neat-water and saturated systems reflecting the surface mobility of the \suldiox~molecules in the two systems.
